---
title: "HW7_MandMs"
---

In this assignment, you'll be picking up where you left off with the M&Ms activity. You'll load the data and perform some basic statistical analysis. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Load the M&Ms dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv

```{r}
mandm <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv")
```

### Task 3: Sample 5 rows of data from the dataset, and then sample 10 rows, saving each to separate dataframes called sample5 and sample10.

```{r}

sample5 <- mandm |> sample_n(5)
sample10 <- mandm |> sample_n(10)

view(sample10)
view(sample5)

```

### Task 4: Calculate the mean and standard deviation of the samples, using one of the colors.

```{r}
sample5 |> summarise(mean(red), sd(red))
  
  sample10 |> summarise(mean(red), sd(red))

```


### Task 5: Calculate the mean and standard deviation of the population, using the same color as Task 4

```{r}

mandm |> summarise(mean(red), sd(red))

```

### Task 6: Compare the sample and population statistics

Compare the three sets of means and standard deviations. Are they similar?

The population and sample10 mean and median for red are more similar than the sample of 
five. In population, the mean was 72.98 and SD 16.45. For the sample of 10, the mean
was 71 and SD was 18. However, there was more variation in the results for the 
sample of 5 with SD being 13 and mean 68.6. Sample 5 wasn't super far off from the 
other two samples, but it was a substantial amount.


### Task 7: Describe the results

Write a few sentences about what you found in the data. What does this tell you about the M&Ms dataset and about the act of random sampling?

The M and M dataset told me that you can really get a good sense of where the
population stands without having to study the entire group. With the correct 
amount in a sample that's sufficiently randomized, you can get the same results
without so much effort or resources. This exercise made me understand why sampling a small
population isn't as shady as one thought. 



When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
