---
title: "Maryland Public School Absenteeism Project"
Name: "Stella S. Canino"
---
 
Introduction:
 In recent years, the rate of absenteeism and chronic absenteeism has increased 
 in the United States. Maryland is not an exemption. The purpose of this 
 statistical analysis is to figure out how absenteeism and chronic absenteeism
 is looking  in the state by viewing 2019-2024 school year data
 provided from the Maryland Department of Education database.
 Each of these datasets contains attendance and absence totals and percentages for every individual school
 in the county, for the entire district, and for the district broken down by elementary, middle, and high school levels.

 We will see who are the schools with the highest levels of absenteeism from 
 this data set and determine their characteristics. 
 
 Step 1: Input of data and cleaning
 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
```

# I am loading the libraries and absenteeism datasets. 

```{r}
library(tidyverse)
library(janitor)
library(knitr)      # For nice tables
library(lubridate) 
```

```{r}
absenteeism_2024 <- read_csv("~/Documents/GitHub/JOUR405_SSCQ/project/Attendance_2024.csv") #Post-pandemic
absenteeism_2023 <- read_csv("~/Documents/GitHub/JOUR405_SSCQ/project/Attendance_2023.csv") #Post-pandemic
absenteeism_2022 <- read_csv("~/Documents/GitHub/JOUR405_SSCQ/project/Attendance_2022.csv") #Pandemic
absenteeism_2021 <- read_csv("~/Documents/GitHub/JOUR405_SSCQ/project/Attendance_2021.csv") #Pandemic
absenteeism_2020 <- read_csv("~/Documents/GitHub/JOUR405_SSCQ/project/Attendance_2020.csv") #Pandemic
absenteeism_2019 <- read_csv("~/Documents/GitHub/JOUR405_SSCQ/project/Attendance_2019.csv") #Pre-pandemic
```

#Here I'm viewing the data to determines the columns, rows and outputs presented.

```{r}
glimpse(absenteeism_2024)
glimpse(absenteeism_2023)
glimpse(absenteeism_2022)
glimpse(absenteeism_2021)
glimpse(absenteeism_2020)
glimpse(absenteeism_2019)

#All of the data here has academic year as an interger, and the rest of the numbers as characters. I will be doing the change for this below. 
```

#Here I'm standarizing the names of the columns for all the datasets so it's 
easier for me to type it out and all have the same format.

```{r} 
absenteeism_2024 <- absenteeism_2024 |> clean_names()
absenteeism_2023 <- absenteeism_2023 |> clean_names()
absenteeism_2022 <- absenteeism_2022 |> clean_names()
absenteeism_2021 <- absenteeism_2021 |> clean_names()
absenteeism_2020 <- absenteeism_2020 |> clean_names()
absenteeism_2019 <- absenteeism_2019 |> clean_names()
```

# When I downloaded the 6 datasets values that are supposed to be numeric 
ended up being viewed as characters. I proceeded to mutate the corresponding
columns to numerical. This method of using the function across. I learned
in JOUR472 last semester. 

```{r}

absenteeism_2024 <- absenteeism_2024 |> 
  mutate(across(
    c(attend_rate_pct, days_attended_cnt, days_member_cnt,
      fewer_5_pct, fewer_5_cnt, more_20_pct, more_20_cnt,
      more_90_member_cnt, chronic_absentee_cnt, chronic_absentee_denom, chronic_absentee_pct
    ),
    as.numeric
  ))

absenteeism_2023 <- absenteeism_2023 |> 
  mutate(across(
    c(attend_rate_pct, days_attended_cnt, days_member_cnt,
      fewer_5_pct, fewer_5_cnt, more_20_pct,more_20_cnt,
      more_90_member_cnt, chronic_absentee_cnt, chronic_absentee_denom, chronic_absentee_pct
    ),
    as.numeric
  ))

absenteeism_2022 <- absenteeism_2022 |> 
  mutate(across(
    c( attend_rate_pct, days_attended_cnt, days_member_cnt,
      fewer_5_pct, fewer_5_cnt, more_20_pct,more_20_cnt,
      more_90_member_cnt, chronic_absentee_cnt, chronic_absentee_denom, chronic_absentee_pct ),
    as.numeric))

absenteeism_2021 <- absenteeism_2021 |> 
  mutate(across(
    c( attend_rate_pct, days_attended_cnt, days_member_cnt,
      fewer_5_pct, fewer_5_cnt, more_20_pct,more_20_cnt,
      more_90_member_cnt, chronic_absentee_cnt, chronic_absentee_denom, chronic_absentee_pct),
    as.numeric))

absenteeism_2020 <- absenteeism_2020 |> 
  mutate(across(
    c(attend_rate_pct, days_attended_cnt, days_member_cnt,
      fewer_5_pct, fewer_5_cnt, more_20_pct,more_20_cnt,
      more_90_member_cnt, chronic_absentee_cnt, chronic_absentee_denom, chronic_absentee_pct
    ), as.numeric))

absenteeism_2019 <- absenteeism_2019 |> 
  mutate(across(
    c(
      attend_rate_pct, days_attended_cnt, days_member_cnt,
      fewer_5_pct, fewer_5_cnt, more_20_pct,more_20_cnt,
      more_90_member_cnt, chronic_absentee_cnt, chronic_absentee_denom, chronic_absentee_pct
    ),
    as.numeric
  ))
```

Now that I have transformed my columns to numerical ones. I need to filter
out values that don't have numbers. I do this because if I don't it might
jam or make difficult building statistical models and analysis in R. 
I used the filter function.

```{r}
absenteeism_2024 <- absenteeism_2024 |> filter(days_attended_cnt != "*")

absenteeism_2023 <- absenteeism_2023 |> filter(days_attended_cnt != "*")
                                               
absenteeism_2022 <- absenteeism_2022 |> filter(days_attended_cnt != "*")                                           
absenteeism_2021 <- absenteeism_2021 |> filter(days_attended_cnt != "*")

absenteeism_2020 <- absenteeism_2020 |> filter(days_attended_cnt != "*")

absenteeism_2019 <- absenteeism_2019 |> filter(days_attended_cnt != "*")
```

These are columns in my 6 datasets that don't provide any important information 
to my analysis. I decided that eliminating from my analysis in R would help me
be more time-efficient with my analysis. I do this by selecting create_date and 
lea_number and using the - to tell R I don't wanna see this no more. 

```{r}
absenteeism_2024 <- absenteeism_2024 |>
  select(-create_date, -lea_number)

absenteeism_2023 <- absenteeism_2023 |>
  select(-create_date, -lea_number)

absenteeism_2022 <- absenteeism_2022 |>
  select(-create_date, -lea_number)

absenteeism_2021 <- absenteeism_2021 |>
  select(-create_date, -lss_number)

absenteeism_2020 <- absenteeism_2020 |>
  select(-create_date, -lss_number)

absenteeism_2019 <- absenteeism_2019 |>
  select(-create_date, -lss_number)
```

#In my data set LEA and LSS is the name of the county/school district an individual
school belongs too. The name changed and means the same thing in all the datasets.
I decided to make my columns more consistent by renaming. 

```{r}
#CHANGED SO ITS MORE CONCISE. 
absenteeism_2020 <- absenteeism_2020 |>
  rename(lea_name = lss_name)

absenteeism_2021 <- absenteeism_2021 |>
  rename(lea_name = lss_name)

absenteeism_2019 <- absenteeism_2019 |> 
  rename(lea_name = lss_name)
```

**STATISTICAL ANALYSIS SECTION**

Initial Analysis. Before performing statistical tests like ANOVA and
Z-Score, I wanted to make initial calculations on my datasets separately to get
an idea on what the outputs and results are being manifested by viewing data
from 2019 (pre-pandemic) and 2024 (post-pandemic). 

This is the 2019 and 2024 absenteeism dataset. I filtered A and All. A represents
the entire aggregate of the school district itself and also by school type
(elementary, middle, and high school). A is different because it specifically 
targets schools that has all three school types or 2. I did this because
I wanted to focus every school individually by type. Having all the 
aggregate data would have given me a false result. 

I also analyzed this separately to see what school districts have the largest
chronic absentee percentage.

#1A
Important Result: The school districts with the highest rate of chronic
absenteeism in Maryland are: Baltimore City High Schools (60.8%), All Baltimore City Schools (42.5%),
All Cecil High Schools (35.9%), All Baltimore City Elementary Schools	(35.1%),
and All Baltimore City High Schools (34.1%)

Conclusion: Baltimore City tops the list for most chronic absenteeism. 

```{r}
#1A
absenteeism_2019 |> filter(school_number == "A") |>
  arrange((desc(chronic_absentee_pct)))

```
#2A

The school with the highest amount of chronic absenteeism in 2019 is 
Renaissance Academy in Baltimore City. Then its Frederick Douglass High in Baltimore City and Green Valley Academy at Edgar Allan Poe (high school and middle school) in PG County. In the lists, it tends to be Baltimore City that tops the list with PG County and Baltimore County following.

```{r}
#2A
absenteeism_2019 |> filter(school_number != "A", school_type != "All Students") |> 
  arrange(desc(chronic_absentee_pct))

#I did this one just to see the schools that have the highest attendance rate.
#I just wanted to see
absenteeism_2019 |> filter(school_number != "A", school_type != "All Students") |>
  arrange(desc(attend_rate_pct))
           
```
#3A

In school district Baltimore City tops the list for most chronic absenteeism
with 48.7%, Dorchester (38.3%), Allegany (36.4) and Somerset (33.5). 

In 2024, Achievement Academy at Harbor City High, Joseph C. Briscoe Academy,	Meadowood Education Center are the schools with the highest rates of absenteeism.
All are in Baltimore City.

```{r}
#This is my code to view what are schools in Maryland with the highest rate of chronic absenteeism. 

#3A
absenteeism_2024 |> filter(school_number != "A", school_type != "All") |> 
  arrange(desc(chronic_absentee_pct))

#I did this one just to see the schools that have the highest attendance rate.
#I just wanted to see
absenteeism_2024 |> filter(school_number != "A", school_type != "All") |>
  arrange((attend_rate_pct))

absenteeism_2024 |> filter(school_number == "A", school_type == "All") |>
  arrange((desc(chronic_absentee_pct)))

```

Step 3: Standard Deviation and Mean of the entire dataset

I combined all of the datasets to bind rows so I can make analysis 
with all the data collected about the general relationship with absenteeism
in all the 6 school years.

```{r}
#combinar la data
#LOOK AT JOUR472 HW 4 or 5??

total <- bind_rows(absenteeism_2024, absenteeism_2023, absenteeism_2022, absenteeism_2021, absenteeism_2020, absenteeism_2019)

glimpse(total)
```

I wanted to see what is the average and standard deviation of the percentage
of the chronic absenteeism. The results I received is a mean of 33.42 and 
an SD of 17.42. 

```{r}
##Look for total mean, standard deviation. Then also check for the year-by-year one. 

#I'm looking at the chronic absenteeism percentage
#How to do it with columns is in JOUR472 github???? 
total |> 
  summarise(
    mean_chronic_absent = mean(chronic_absentee_pct, na.rm = TRUE),
    sd_chronic_absent = sd(chronic_absentee_pct, na.rm = TRUE)
  )
```

Because a 33% mean is a relatively small number, I wanted to view the individual school years average to see if there are differences with some years going up 
and some going down. 

Results: It seems that chronic absenteeism increased from 2021 and 2024, but is
starting to recover back to pre-pandemic levels.

```{r}
total |> 
  filter(academic_year == "2019") |> 
  summarise(
    mean_chronic_absent = mean(chronic_absentee_pct, na.rm = TRUE),
    sd_chronic_absent = sd(chronic_absentee_pct, na.rm = TRUE)
  )
```

```{r}
total |> 
  filter(academic_year == "2020") |> 
  summarise(
    mean_chronic_absent = mean(chronic_absentee_pct, na.rm = TRUE),
    sd_chronic_absent = sd(chronic_absentee_pct, na.rm = TRUE)
  )
```

```{r}
total |> 
  filter(academic_year == "2021") |> 
  summarise(
    mean_chronic_absent = mean(chronic_absentee_pct, na.rm = TRUE),
    sd_chronic_absent = sd(chronic_absentee_pct, na.rm = TRUE)
  )
```

```{r}
total |> 
  filter(academic_year == "2022") |> 
  summarise(
    mean_chronic_absent = mean(chronic_absentee_pct, na.rm = TRUE),
    sd_chronic_absent = sd(chronic_absentee_pct, na.rm = TRUE)
  )
```

```{r}
total |> 
  filter(academic_year == "2023") |> 
  summarise(
    mean_chronic_absent = mean(chronic_absentee_pct, na.rm = TRUE),
    sd_chronic_absent = sd(chronic_absentee_pct, na.rm = TRUE)
  )
```

```{r}
total |> 
  filter(academic_year == "2024") |> 
  summarise(
    mean_chronic_absent = mean(chronic_absentee_pct, na.rm = TRUE),
    sd_chronic_absent = sd(chronic_absentee_pct, na.rm = TRUE)
  )
```

STEP 4: Z-SCORE TEST

I'm using the Z-Score test because it will help me determine if there is a 
significant difference between the general mean of my population and the chronic
absentee percentage of each school. This will help me find the outliers of my 
dataset, which means it's the schools that have the highest variation from the
mean. The outliers will be the schools with the largest chronic absentees. 

This is the general Z-score test. When you open the chart now in the open
environment. You see the Z-Score of every single school in the past 6 school years.

```{r}
total <- total |>
  mutate(
    mean_chronic_absent = mean(chronic_absentee_pct, na.rm = TRUE),
    sd_chronic_absent = sd(chronic_absentee_pct, na.rm = TRUE),
    z_score = (chronic_absentee_pct - mean_chronic_absent) / sd_chronic_absent
  )
```

Now, I want to see the outliers and extreme outliers of my dataset. This is to view
which are the schools that have the highest rate of chronic absenteeism compared
to the rest of the dataset. The point of my analysis is to find the schools
with high number of absenteeism. 

There are 394 outliers from all 6 school years in the total dataset. This is from the range of 2-3 Z-scores. My main focus is the extreme outliers, but to get to that I have to first find the general outliers. 

```{r}
outlier <- total |>
  filter(school_number != "A", school_type != "All") |>
  filter(abs(z_score) > 2) |>
  select(academic_year, school_name, school_type, z_score) |>
  arrange(desc(z_score))
```

This is the Z-score extreme outliers test. 

There are 71 extreme outliers in the entire data set. The school that are the biggest outliers are Northeast EDLP at Parkville High School in 2021, Meadowood Education Center in 2021 and 2023, Renaissance Academy in 2022 and Green Valley Academy at Edgar Allan Poe in 2021. 

The majority are from Baltimore City and Baltimore County.

```{r}
extreme_outliers <- total |>
  filter(school_number != "A", school_type != "All") |>
  filter(abs(z_score) > 3) |>
  arrange(desc(abs(z_score)))

```

**ANOVA TEST**

We have established the Z-scores of the entire dataset. We were able to identify
schools with the highest rates of chronic absenteeism in different years.

Now, I will be using the ANOVA Test to compare the mean of the three school types in Maryland
to determine the difference between them. This will help
me define which school type has the highest rate of absenteeism. 

Part 1: 
First thing that I will do is create a new dataset that has all the averages
from all the schools for their chronic absenteeism. Then I grouped it by school name, county, and school type.

I filtered out "All" and "All Students" because that is the aggregation of the
entire school district and school district type and I didn't want those outputs
to change my results. It would have inflated my calculations.

```{r}
school_averages <- total |>
  group_by(school_name, lea_name, school_type) |>
  summarize(
    avg_chronic_absent = mean(chronic_absentee_pct, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(
    !is.na(avg_chronic_absent),
    !school_type %in% c("All", "All Students")
  )
```

Part 2: The test

This is my ANOVA Test. I wanted to see the relationship between the average chronic absenteeism and also the school type.

The results show that there is a very low probability that the differences between school types occurred by random chance (p < 2e-16). There is a statistically significant difference in chronic absenteeism rates across school types.

The F-value of 78.69 suggests that the variation between school types is large compared to the variation within each type, meaning school type is a meaningful factor.

However, school type does not fully explain the differences in absenteeism. There is 375101 residuals, meaning that most variation is due to other factors not included in the model. 

Factors like issues with transportation and wealth inequality could be factors that explain the variation between schools. 

```{r}
period_anova <- aov(avg_chronic_absent ~ school_type, data = school_averages)
summary(period_anova)
```
Part 3: TUKEY TEST

Once my ANOVA test proved that the difference between school-type and chronic absenteeism is statistically significant, I will be using the Tukey HSD Test.

This will help me compare the difference of the means grouped by ANOVA, and will let me identify which of the the 3 groups are significantly different from each other.

Results:

High schools have on average 11.86 percentage points higher chronic absenteeism than elementary schools. This is statistically significant.cMiddle schools also have higher chronic absenteeism than elementary schools — by about 5.14 percentage points.

This pattern is statistically significant across all pairwise comparisons. I have solid evidence that chronic absenteeism increases with school level.

```{r}
tukey_results <- TukeyHSD(period_anova)
print(tukey_results)
```
**Visualization - Boxplot**

To visualize my ANOVA test results, I created a box plot that shows the 
difference between the averages of the three school types (elementary,
middle and high school). The overall trend shows that high schools have the
highest average rate of chronic absenteeism. We can also see that high school
has the higher variability. 

All three school types have outliers that are more than 50%. Although, elementary schools have the lowest average chronic absenteeism rate compared to the other two,
it has the largest number of outliers meaning that a good amount of schools have 
schools with really high rates. 

I found the code to creating a boxplot from JOUR472 last semester.

```{r}
# Find code in JOUR472 Github and in one of the anova test notebook. 
ggplot(school_averages, aes(x = school_type, y = avg_chronic_absent, fill = school_type)) +
  geom_boxplot(alpha = 0.9) +
  scale_fill_manual(values = c("purple", "pink", "magenta")) + #THIS IS FROM JOUR472 notebook 
  labs(
    title = "Chronic Absenteeism by School Type in Maryland Public Schools",
    x = "",
    y = "Average Chronic Absenteeism Rate (%)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```

This code is to see what are the schools with the highest average chronic absenteeism. 

```{r}
school_averages |> arrange(desc(avg_chronic_absent))
```
**Story Pitch** 

I want to write a data-driven story about how chronic absenteeism affects mostly high school
students in Maryland and which are the schools struggling the most and who are the students
in these schools that skip class. 

In my statistical analysis, I found that chronic absenteeism in Maryland public schools averaged
33.42% and a standard deviation of 17.42% from 2019 to 2024. When I calculated, the 
average of each public school using the total dataset, I found the top schools for
chronic absenteeism. Among the top 10 schools all had over 85% absentee rates and student
populations that were over 97% minority. Most were located in Baltimore City.

With this information, I would go to the schools to view the conditions and interview
students, teachers, principals and other personnel about why they have such high rates
 of absenteeism. 

Using our Z-score analysis, I identified 71 schools as outliers over six school years. I plan to visit also visit these schools to investigate and explore what strategies, if any, were used to lower their chronic absenteeism from their peak established in the analysis. 

From my Tukey test, we found that chronic absenteeism increases with school type, with high schools having 11.86% more average chronic absenteeism than elementary schools and middle schools are 6.72 points lower than high schools. 

I will interview officials in the Maryland Department of Education to figure out what is being implemented statewide to tackle this issue. Furthermore, I would interview community leaders and parents in the area to have their input on why students aren't going to class. 

From my ANOVA test, I was able to find that there is statistically significant
difference between the averages between the 3 school types. However, the ANOVA
model results about school type only explains a small percentage of variation.


With that in mind, I will have to analyze additional data relating to transportation, income inequality, etc to be able to inform readers of possible reasons why these schools have such high levels of absenteeism.

Using the dataset for ANOVA and Tukey, I would include my box plot in my reporting to help visualize
this to readers. Although high schoolers have the higher rates of chronic absenteeism, middle schools and
elementary schools are also facing this crisis as well. I would make a point that elementary schools has the lowest average of chronic absenteeism, but also has the highest number of outliers that range from 50% to almost 80%.

With all of these elements together, I wil be able to write an in-depth investigative story about chronic absenteeism in Maryland.








